Visualizing Conditional Dependencies

### 2018-02-07 paper/sections/section1.tex
submitted paper

\subsection{Categorical}

\subsubsection{Cross Tabulation Visualization}
\begin{figure}
\includegraphics[width=1\textwidth]{swsg.png}
\caption{Graphic designer Tim Leong visualizes a cross tabulation of the severity of a situation against the language used in each situation to show the relative frequency of really bad situations in Star Wars. The ordering of the phrases by the order in which the movies came out and the slight semantic variations between each phrase can be further mined for potential conditional dependencies between these underlying aspects of the phrase, the phrase itself, and the severity of the situation. Chart is from Star Wars Super Graphic \cite{leong_star_2017}}
\label{fig:starwars}
\end{figure}

Categorical data can be likened to set measurement, where each category is a set
and researchers often want to know how many members each set has and what
intersections are common between sets \cite{agresti_categorical_2011,schneider_set-theoretic_2012}. In exploring these intersections, the question arises of which categories appear in conjunction with each other and if any one category is the driving force underlying the other categories. For example in frequent item set detection, the task is explicitly to figure out what is the minimum set of items that can be used to predict the occurrence of other items in the basket \cite{leskovec_mining_2014,Srikant:1997:MAR:3001392.3001404} The most common approach is to create a list of frequencies, sometimes of sets of one category or a cross-table\cite{goodman_measures_1991} which is an aggregate function applied to the product of two or more sets. A visual example of a simple cross tabulation can be see in figure~\ref{fig:starwars}, wherein designer Tim Leong looks at the probability of a situation being bad given the structure and content of a given phrase\cite{leong_star_2017}. Each phrase acts as set of size 1, the cross tabulation in this case being the likelihood of the situation being bad. Because the phrases are ordered in temporal order, Leong's visualization suggests that there's a higher likelihood of bad events happening near the middle of the time-line, indicating that severity of situational badness and the type of phrasing used could be conditionally dependent on the location of the story arc.

\begin{figure}
\includegraphics[width=1\textwidth]{hanging_rootgram.png}
\caption{The hanging rootogram moves the histogram bars so that they meet the expected values of the density estimator. This in turn means that deviations between the expected and observed frequencies can be measured more easily as distance between the bottom of the rectangle to a horizontal 0, rather than as a difference between the top of the rectangle and a point on a distribution curve.}
\label{fig:hanging_rootogram}
\end{figure}

 While a table is encouraged for a small number of variables
 \cite{munzner_what_2014}), bar graphs and histograms are typically used to display differences between subsets \cite{ioannidis_history_2003-1, friendly_brief_2006}. Variations on histograms  include the hanging rootogram shown in figure~\ref{fig:hanging_rootogram} and other distribution oriented plots \cite{tukey_exploratory_1977, friendly_visualizing_2000}. While conditional relationships can somewhat be visualized using other categorical visualizations such as fourfold displays, mosaic displays, and logit regression models, these plots are somewhat limited to 3 sets of categories
 \cite{friendly_visualizing_2000}.

 \subsubsection{Set Visualization}
\begin{figure}
  \includegraphics[width=\textwidth]{upsetfig1}
   \caption{UpSet dashboard displaying the relationships amongst movie genres. The left hand side lets the user query the database for a set of movies, listed in the set menu. It then shows how these movies intersect with the other genres of sets. The middle panel shows how many movies are in each intersecting set, and the distribution of select quantitative variables. The right hand panel breaks out the movies in one of these subsets into a scatter plot and a list. This combined visualization tool allows a user to see if quantitative attributes like ratings or times watched can be conditionally dependent on the genre of the movie. This figure comes from UpSet: Visualization of Intersecting Sets \cite{lex_upset_2014}}
   \label{fig:upsetfig}
\end{figure}

The UpSet tool is designed to display a much larger number of set intersections, frequency information about set membership, and the aggregate information of quantitative attributes filtered on sets\cite{lex_upset_2014}. As shown in figure~\ref{fig:upsetfig}, the UpSet tool provides linked set and element views. The set view provides the intersections and their
aggregates, the frequencies of each set and intersection, and the aggregate
statistics of the associated quantitative variables, while the element view shows
selected elements and information about their set membership. The set view yields the distributions of each of the attributes. For example the selected row (in the gray box) are the distributions of each quantitative variable in the comedy set. Every row underneath is the distribution of that quantitative variable restricted to movies in the intersection of the selected sets. For example, the distribution of ratings of movies that are both comedies and dramas. The Upset tool
also facilitates comparison of filtered sets via scatter-plot, which acts as a
visualization of the attributes of the data conditioned on membership in the
filtered set.  The UpSet tool provides a window into conditional dependency because it allows for filtering on sets and then updates the scatter and quantile views; like the scatter matrix, UpSet can isolate a genre or filter on ranges of quantitative variables such that the user can try a variety of approachs to see if there is directionality in the co-occuring events. 
\end{document}

>>>>

\subsection{Categorical}
\label{sec:categorical}
Categorical data, including vectors of categorical observations, can be represented as sets; researchers then query set cardinality, membership, and intersections \cite{agresti_categorical_2011,schneider_set-theoretic_2012}. In exploring these intersections, the question arises of which categories appear in conjunction with each other and if any one category is the driving force underlying the frequency of other categories. For example, in frequent item set detection, the task is to figure out the minimum set of items that can be used to predict the occurrence of other items in the set \cite{leskovec_mining_2014,Srikant:1997:MAR:3001392.3001404}. By representing categorical variables as sets, we can exploit methods of set visualization to explore dependencies amongst categorical variables.

\subsubsection{Cross Table}
\label{sec:crosstab}
\begin{figure}[H]
\includegraphics[width=1\textwidth]{swsg.png}
\caption{Graphic designer Tim Leong visualizes a cross table of a phrase against the severity of the situation in which it was said. Because the situations are ordered chronologically, this illustration shows that the most severe situations occur in the middle of the Star Wars' timeline. Chart is from \textit{Star Wars Super Graphic} \cite{leong_star_2017}}
\label{fig:starwars}
\end{figure}

One of the most common approaches for understanding relationships between categorical variables is to aggregate a function applied to the product of two or more sets; this computation is called a cross table \cite{goodman_measures_1991}. Figure~\ref{fig:starwars} is a visual example of a cross table where the sets are a phrase spoken by a character in a Star Wars film and the severity of the situation in which the phrase is spoken \cite{leong_star_2017}. Designer Tim Leong structures the visualization such that the phrases are chronologically ordered with respect to the Star Wars' timeline and he lines up the phrases such that variations are readily apparent. He then encodes the probability of danger as the height of the yellow bar such that the more dangerous, the more it encroaches on the \textit{Really Bad Situation} side of the visualization. Each phrase is a set with one element (the phrase), and the cross tabulation is the likelihood that the speaker is in a dangerous situation. Because the phrases are ordered in temporal order, Leong's visualization suggests that there is a higher likelihood of bad events happening near the middle of the timeline, indicating that severity of situational badness and the type of phrasing used could be conditionally dependent on the location of the event- in the story arc. Leong effectively uses a cross table to visualize a small number of groups (phrases) and a very small number of variables (phrases, situations). Cross-tabulations shine when there are a few variables and a small number of groups, but get very unwieldy as the size of the set product grows.

\begin{figure}[H]
\includegraphics[width=1\textwidth]{hanging_rootgram.png}
\caption{The hanging rootogram shifts the histogram bars up so that they intersect with the expected values of the density estimator. This means that deviations between the expected and observed frequencies can be measured as the distance between the bottom of the rectangle and horizontal line at x=0.}
\label{fig:hanging_rootogram}
\end{figure}
 While a simple table is encouraged for a small number of variables or categories \cite{munzner_visualization_2014}), bar graphs and histograms \cite{ioannidis_history_2003-1, friendly_brief_2006} are more appropriate when the variable has a large number of unique categories or the dataset has a large number of variables. Variations on histograms  include the hanging rootogram shown in figure~\ref{fig:hanging_rootogram} and other distribution oriented plots \cite{tukey_exploratory_1977, friendly_visualizing_2000}. While conditional relationships can be visualized using other categorical visualizations such as fourfold displays, mosaic displays, and logit regression models, these plots are somewhat limited to 3 categorical variables \cite{friendly_visualizing_2000}. For datasets with more complex categorical relationships, we can look to set oriented visualization tools

 \subsubsection{Set Intersection}
 \label{sec:setintersection}
\begin{figure}[H]
  \includegraphics[width=\textwidth]{upsetfig1}
   \caption{UpSet dashboard displaying the relationships between a movie's genre and its release date, ratings, and popularity. The left panel of the dashboard lets the user query the database for all movies in a selected genre. It then shows which other genres these movies belong to. The middle panel shows the distribution of release dates, ratings, and popularity of the movies in each intersecting set of genres. The right panel lists the individual movies in one of these intersecting sets and plots their quantitative attributes as scatter plot. This combined visualization tool allows a user to see if quantitative attributes like ratings or popularity are conditionally dependent on the genre. This figure comes from \textit{UpSet: Visualization of Intersecting Sets} \cite{lex_upset_2014}}
   \label{fig:upsetfig}
\end{figure}

The UpSet tool is designed to display a large number of set intersections, frequency information about set membership, and information about the distribution of quantitative attributes of elements in a set \cite{lex_upset_2014}. As shown in figure~\ref{fig:upsetfig}, the UpSet tool provides linked set and element views so that users can get a top level view of the characteristics of a genre of movies and also information about the movies in that genre. The set view on the left panel of the dashboard lets users select a genre and then displays information about how many movies are in that genre and what other genres those movies belong to. The middle panel shows the distribution of the release dates, average ratings, and popularity of the movies in each of the sets shown on the right side. The right panel element view displays the movies in the selected genre and the scatter plot shows each movie's ratings and popularity. The UpSet tool provides a window into conditional dependency because a user can see if attributes of the movies in a genre, such as average movie rating and release date, are conditionally dependent on genre. Users can select a genre (e.g. comedy) and look at how the attributes of the movies in that set distribute and then compare those results against the distributions of other genres. While the UpSet tool can handle data that is too complex for cross tables (\ref{sec:crosstab}), it can not preserve structure and is limited to showing intersections of sets of one variable type (e.g. only genres) and a low number of quantitative variables.

\end{document}

### 2018-02-07 paper/sections/section2.tex
submitted paper

\subsubsection{Radial Visualizations}
\begin{figure}
    \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.2]{cities.png}
    \caption{The size and shape of each polygon acts as an indicator of Internet usage over time. All events in all cities seem to have expanded over time, indicating a conditional dependence on time.}
    \label{fig:radarchart}
    \end{subfigure}

    \bigskip
    \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.2]{flower.png}
    \caption{The size of each flower indicates Internet usage. The grid arrangement in this chart facilitates comparison both across time (horizontally) and city (vertically). As in figure~\ref{fig:radarchart}, Internet usage seems to conditionally dependent on time.}
    \label{fig:flowerchart}
    \end{subfigure}

    \bigskip
    \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.2]{circle.png}
    \caption{The circle chart prioritizes intensity of each type of usage, most clearly showing the popularity of shopping and SMS. This visualization facilitates testing a hypothesis of a conditional dependency amongst the Internet usage variables by isolating the similar pairwise color patterns in the presence and absence of a third usage type.}
    \label{fig:circlechart}
    \end{subfigure}
\caption{All three visualizations show Internet usage in 5 cities, from 2003 to 2012. Images are from Off the Radar: Comparative Evaluation of Radial Visualization Solutions for Composite Indicators \cite{albo_off_2016}}
\label{fig:radial}
\end{figure}
Radial visualizations are designed to show composite indicators (CI)\cite{chambers_graphical_1983}; a CI is a type of visualization wherein the multivariate observation is  mapped to a single visual idiom. For example, the whole line across all the Y-axis in a parallel coordinates plot (PCP) is a CI. A radar plot, as shown in figure~\ref{fig:radarchart}, is just a PCP where the two sides of the axis meet. Like a PCP, the radar chart shows pairwise relationships between variables that share nodes, but the radial shape also means it can show pairwise relationships between all variables. In 2010 in figure~\ref{fig:radarchart}, the more rectangular polygon for city 2 relative to the somewhat trapezoid polygon for city 4 indicate that Internet usage is characteristically different for the two cities. The conditional dependency factors in because the radial plot shows changes over time, by city, across multiple components of Internet usage. The flower chart in figure~\ref{fig:flowerchart} is very similar to the radar chart, also highlighting a potential dependency on time due to the use of small multiples and the flowers progressively getting larger. The circle chart in figure~\ref{fig:circlechart} stands out particularly because time progressions are not as readily visible, but instead this chart, especially with a finer grained colormap, best highlights the specific usage variables. This chart most clearly shows that shopping and sms seems to happen in tandem, and that clear color delineation also makes it easier for a user to scan the other variables to see how frequently they occur given the relationship between shopping and SMS.

\subsubsection{Cluster Based Visualizations}
\begin{figure}
\includegraphics[width=1\textwidth]{calendar.png}
\caption{Figure from \cite{van_wijk_cluster_1999}. Van Wijk and Van Selow visualized the number of employees present at a research facility. This visualization shows that daily attendance seems to be more strongly dependent on day of week than time of day or day of the year. Figure is from Cluster and Calendar based Visualization of Time Series Data \cite{van_wijk_cluster_1999}}
\label{fig:calendar}
\end{figure}

Finding patterns and trends in large datasets, whether univariate or multivariate, is often a matter of trying to find representational fingerprints of common or outlying events in the dataset. This is frequently done through the use of a clustering algorithm. Figure~\ref{fig:calendar} shows an analysis of employee attendance at a research facility in the Netherlands\cite{van_wijk_cluster_1999}. Because the individual observations were recorded every 6 hours for a year, van Wilk and van Selow transformed the hourly observations into a matrix where each vector holds the observations for a given day. They then used an agglomerative clustering algorithm \cite{kaufman_agglomerative_1990} to find representative days, and color matched these clusters to a calendar of days to visualize how these clusters were distributed over time.  In this study, time is separated into three classes of events: time of day, day of week, and date. The researchers found that both the shape and magnitude of the attendance curve changed relative to the day of week, indicating a conditionally dependent relationship between attendance and day of week. 
\end{document}

>>>>

Unlike in section~\ref{sec:categorical}, some datasets have a very large number of observations or the observations are a vector of quantitative measurements that work in tandem. One approach for handling these types of datasets is to develop categorical representations of these quantitative variables.

\subsubsection{Radial Charts}
\label{sec:radial}
One method of transforming quantitative vectors into categorical representations is to visualize the vector using a radial graph. Radial visualizations are designed to show composite indicators (CI) \cite{chambers_graphical_1983}; a CI is a type of visualization where the multivariate observation is mapped to a single visual idiom. For example, the line representing the observation vector in a parallel coordinates plot is a CI. 
\begin{figure}[H]
    \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.2]{cities.png}
    \caption{The size and shape of each polygon acts as an indicator of Internet usage over time.}
    \label{fig:radarchart}
    \end{subfigure}

    \bigskip
    \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[scale=0.15]{flower.png}
    \caption{The size of each flower indicates Internet usage. The grid arrangement in this chart facilitates comparison both across time (horizontally) and city (vertically).}
    \label{fig:flowerchart}
    \end{subfigure}

    \bigskip
    \begin{subfigure}{\textwidth}
    \centering
    \includegraphics[width=1\linewidth]{circle.png}
    \caption{The circle chart prioritizes intensity of each type of usage, most clearly showing the popularity of shopping and SMS. This visualization facilitates testing a hypothesis of a conditional dependency amongst the Internet usage variables by isolating similar pairwise color patterns in the presence and absence of a third usage type.}
    \label{fig:circlechart}
    \end{subfigure}
\caption{Internet usage in 5 cities from 2003 to 2012. Since all types of Internet usage in all the cities seem to have expanded over time, this suggests a conditional dependence of different types of usage on time. Images are from \textit{Off the Radar: Comparative Evaluation of Radial Visualization Solutions for Composite Indicators} \cite{albo_off_2016}}
\label{fig:radial}
\end{figure}
A radar plot, as shown in figure~\ref{fig:radarchart}, is a parallel coordinate plot (PCP) where the ends of the line representing the observation are joined. Like a PCP, the radar chart shows pairwise relationships between variables that share nodes, but the radial shape also means that the polygon it forms encodes information about the relationship between all the variables. In figure~\ref{fig:radarchart} in 2010, the more rectangular shape of the city 2 Polygon relative to city 4's trapezoidal polygon signifies a characteristic difference between Internet usage in the two cities. Since figure~\ref{fig:radarchart} is arranged in small multiples of radar charts showing changes over time, by city, and across multiple types of usage, the figure can be used to examine whether this is a conditional dependency on time or city or usage type. Similarly, the flower chart in figure~\ref{fig:flowerchart} also highlights a potential dependency on time due to the flowers progressively getting larger across time. The circle chart in figure~\ref{fig:circlechart} differs from the other charts in figure~\ref{fig:radial} because time progressions are not as readily visible. Instead this chart (especially if the colormap had more steps) most clearly visualizes the frequency of the different types of Internet usage. Figure~\ref{fig:circlechart} shows that shopping and sms seem to happen in tandem, and that clear color delineation also makes it easier for a user to scan the other variables to see how frequently they occur given the relationship between shopping and SMS. The radar, flower, and circle chart visualize Internet usage as circles to varying degrees of success, but radar charts are limited to how much information can reasonably fit on one circle or a small number of circles.

\subsubsection{Clustering}

\label{sec:cluster}
\begin{figure}[H]
\includegraphics[width=1\textwidth]{calendar.png}
\caption{Van Wijk and Van Selow visualized the number of employees present at a research facility. This visualization shows that daily attendance seems to be more strongly dependent on the day of the week than on the time of day or the day of the year. Figure is from \textit{Cluster and Calendar based Visualization of Time Series Data} \cite{van_wijk_cluster_1999}}
\label{fig:calendar}
\end{figure}

If the data is too big for techniques that try to visualize the observations directly, one approach is to instead visualize representational fingerprints of common or outlying events in the dataset. This is frequently done through the use of a clustering algorithm. Figure~\ref{fig:calendar} shows an analysis of employee attendance at a research facility in the Netherlands \cite{van_wijk_cluster_1999}. The individual observations were recorded every 6 hours for a year, so the resulting dataset was a $(\text{day} \times \text{hours})$ matrix of the number of employees at work for that day and time. To reduce the dimensionality of this dataset to something easier to visualize, they used an agglomerative clustering algorithm \cite{kaufman_agglomerative_1990} where the days were observations and the hours were the feature vectors. The results of this clustering were used to create daily attendance curves representative of the days in each cluster; these representative curves are shown in the timeseries plot in figure~\ref{fig:calendar}. They visualized the cluster membership of the days by coloring them the same as their representative cluster, and preserved the temporal structure of the data by presenting this cluster membership as a calendar. The researchers found that both the shape and magnitude of the attendance curve changed relative to the day of the week, indicating a conditionally dependent relationship between attendance and the day of the week. While the visualization Van Wijk and Van Selow built is very specific to the question they were tackling, clustering techniques and the use of visual idioms related to the structure of the data (such as a calendar to present discretized time information) can be readily generalized. The main limitations to applying their approach to other datasets are that clustering inherently reduces dimensions and clustering techniques are limited when the dataset is a mix of categorical and quantitative variables.
\end{document}
